---
title: "Timeline Check In"
author: "Macey Hartmann and Erick Morales Oyola"
date: "22 May 2026"
bibliography: references.bib
format: 
  pdf:
    include-in-header: 
       text: |
         \usepackage{fvextra}
         \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
         \DefineVerbatimEnvironment{OutputCode}{Verbatim}{breaklines,commandchars=\\\{\}}
editor: source
execute: 
  message: false
  warning: false
---

# Setup

## Packages

We loaded packages for data wrangling, file organization, missing-data exploration, and visualization. The main workflow uses `tidyverse` for cleaning and summarizing vegetation data, `here` for file paths, and plotting packages for exploratory figures.

```{r}
#| label: packages
#| echo: false

# general use
library(tidyverse)

# file organization
library(here)

# data visualization
library(patchwork)
library(naniar)
```

## Data 

We used two vegetation datasets. The main dataset, `veg.csv`, contains vegetation survey observations, including species identity, percent cover, cover category, habitat type, transect, site, and year. The metadata file, `vp_veg_metadata.csv`, provides supporting information about vegetation transects, including the number of quadrats sampled within each transect-year.

```{r}
#| label: read-data
#| echo: false

veg <- read_csv(here("data", "veg.csv"))

metadata <- read_csv(here("data", "vp_veg_metadata.csv"))
```

# Ideas for analysis and background 

Our project focuses on salt marsh vegetation at North Campus Open Space (NCOS), with the goal of understanding which plant species characterize this restored wetland habitat and how those patterns have changed across survey years. Rather than treating vegetation cover as only a set of numbers, we are using percent cover to describe how the salt marsh is developing as a plant community over time. In this sense, our project asks how restoration becomes visible through the species that take root, persist, expand, or decline across the marsh.

Salt marsh vegetation is especially useful for understanding restoration trajectory because these plant communities are shaped by strong environmental filters, including salinity, tidal flooding, saturated soils, and seasonal changes in water availability. In southern California salt marshes, native halophytes such as pickleweed (*Salicornia pacifica*) and saltgrass (*Distichlis spicata*) often help define the structure and identity of the habitat [@ferren_1985; @ncos_coastal_salt_marsh]. Because these species are adapted to stressful coastal wetland conditions, changes in their mean percent cover can help indicate whether the restored salt marsh is developing toward a recognizable native marsh community.

This question is particularly relevant at NCOS because the site was restored from the former Ocean Meadows Golf Course into a mosaic of wetland and upland habitats, including salt marsh, brackish marsh, seasonal wetlands, and transitional habitats [@ncos_restoration_plan_2016; @ncos_monitoring_2024]. Broader NCOS monitoring reports evaluate restoration progress across habitat types, while our project narrows in on salt marsh vegetation specifically. By focusing on dominant species and native/non-native cover through time, we can connect the survey data to larger questions about restoration progress, habitat identity, and community development.

Previous work on California wetland restoration has shown that vegetation establishment can be slow, variable, and strongly shaped by local site conditions such as elevation, soil salinity, soil moisture, planting, and adaptive management [@beheshti_2023]. This matters for our analysis because shifts in dominant plant cover are not just visual patterns in a graph. They may reflect ecological processes such as native plant establishment, non-native plant expansion, hydrologic change, disturbance, or interannual variability in rainfall and growing conditions.

## Ecological background

Salt marshes are dynamic coastal wetlands where plant communities exist at the edge of land and water. Their vegetation patterns are shaped by both physical stress and biological interactions, making them useful indicators of environmental conditions and restoration progress. In southern California, seasonal rainfall and dry summers can influence freshwater input, soil moisture, salinity, and plant growth, which may affect how vegetation cover changes from year to year [@ferren_1985].

At NCOS, we expect dominant salt marsh species to include native salt-tolerant plants such as *Salicornia pacifica* and *Distichlis spicata*. The NCOS coastal salt marsh habitat page describes pickleweed as usually the dominant species in coastal salt marshes and also identifies saltgrass, alkali heath, fleshy jaumea, and California saltbush as characteristic salt marsh plants [@ncos_coastal_salt_marsh]. These expectations give us a useful ecological reference point for interpreting our data: if the restored salt marsh is developing toward a recognizable coastal salt marsh community, we would expect these native halophytes to contribute strongly to vegetation cover.

Native and non-native cover are also important to examine because restoration success depends not only on whether plants are present, but on which plants are shaping the habitat. A marsh with high total vegetation cover may still be ecologically different depending on whether that cover is dominated by native salt marsh species or by non-native plants. NCOS monitoring reports describe vegetation cover, native biodiversity, and relative native cover as important indicators of restoration progress [@ncos_monitoring_2024]. Our project builds from that monitoring context by focusing more closely on the species-level patterns that contribute to salt marsh identity through time.

## Planned analyses 

Our first analysis will identify which plant species most strongly characterize salt marsh and closely related brackish marsh vegetation at NCOS. To do this, we will summarize mean percent cover for each plant species across survey years and transects, then rank species from highest to lowest mean cover. This species-rank analysis will help us justify which species we classify as dominant rather than choosing them only from visual inspection.

Our second analysis will examine how the top five dominant species change across survey years. After identifying the five species with the highest mean percent cover across the full filtered dataset, we will summarize their mean percent cover by year and visualize them with a stacked bar plot. In this figure, the y-axis represents the summed mean percent cover of only the top five dominant species, so bar height shows how much cover those dominant species contribute in each year.

Our third analysis will focus specifically on non-native species in the filtered salt marsh/brackish marsh dataset. Rather than only comparing total native and non-native cover, we will identify the five non-native species with the highest mean percent cover across all survey years, then summarize their mean percent cover by year. We will visualize these species with a stacked bar plot so we can see whether non-native cover is being driven by the same species through time or whether different non-native plants become more important in different years. This analysis is useful because a simple native/non-native proportion plot can show the relative amount of non-native cover, but it does not show which species are responsible for that pattern. By focusing on the top non-native species, we can connect non-native cover to actual plant identities and better interpret whether non-native vegetation reflects a consistent community pattern, a temporary increase in a few species, or a shift in the plants contributing to marsh composition.

Our fourth analysis will provide broader context by examining total vegetation cover across survey years. Unlike the dominant-species figure, this analysis includes all native and non-native plant species in the filtered salt marsh/brackish marsh dataset. We will summarize total mean vegetation cover by year and visualize it as a line plot with standard error, which will help us interpret whether changes in species composition are happening alongside broader increases or decreases in overall vegetation cover.

Together, these analyses connect our figures to the main project questions. The species-rank plot identifies the plants that most strongly characterize the habitat, the dominant-species plot shows how the top five species change through time, the proportional native/non-native plot shows how plant origin contributes to habitat identity, and the dominant non-native species plot shows which non-native plants are driving non-native cover across survey years. The total-cover plot then provides broader context for overall vegetation development across the monitoring period.

## Data cleaning and wrangling

We filtered the vegetation dataset to focus on salt marsh and closely related brackish marsh vegetation, then limited the surveys to August and September because late summer and early fall are when vegetation cover is expected to be highest and most comparable across years. We used `complete()` to make implicit absences explicit as zero percent cover, then joined the vegetation data with the metadata file so we could account for the number of quadrats sampled in each transect-year. Finally, we filtered to native and non-native plant cover categories so our analyses focused on vegetation rather than bare ground, thatch, or other non-plant cover types.

```{r}
#| label: missingness
#| echo: true
#| include: false
gg_miss_var(veg)
```

```{r}
#| label: filter-saltmarsh
#| echo: true
saltmarsh <- veg |> 
  # filter to salt marsh habitat only
  filter(str_detect(habitat, "Salt Marsh") | str_detect(habitat, "Brackish Marsh")) |>
  # parse date column and filter to August and September
  mutate(date = as.Date(date),
         month = month(date)) |> 
  filter(month %in% c(8, 9))
```

```{r}
#| label: explicit-zeros
#| echo: true
saltmarsh_complete <- saltmarsh |> 
  complete(year, transect_name, psoc, transect_distance,
           fill = list(percent_cover = 0)) |> 
  group_by(year, transect_name, psoc, cover_category) |> 
  summarize(sum_pc = sum(percent_cover, na.rm = TRUE)) |> 
  ungroup() |>
  unite("year_pool", year, transect_name, remove = FALSE) |> 
  left_join(metadata, by = "year_pool") |>
  select(-year.y, -transect_name.y) |>
  rename(year = year.x,
         transect_name = transect_name.x) |>
  mutate(mean_pc = sum_pc / num_quad)
```

```{r}
#| label: filter-plant-cover
#| echo: true
saltmarsh_plants <- saltmarsh_complete |> 
  filter(cover_category %in% c("NATIVE COVER", "NON-NATIVE COVER"))
```

## Summaries for planned analyses

After cleaning the data, we created summary datasets for each planned figure. We ranked species by mean percent cover, identified the top five dominant species overall, identified the top five non-native species, summarized proportional native and non-native cover by year, and calculated total mean vegetation cover across survey years. These summaries allow us to connect each visualization directly to one of our project questions.

```{r}
#| label: summarize-species-rank
#| echo: true
# summarize-species-rank — keep SE this time
species_rank <- saltmarsh_plants |> 
  group_by(psoc, cover_category) |> 
  summarize(mean_pc = mean(mean_pc, na.rm = TRUE),
            se_pc = sd(mean_pc, na.rm = TRUE) / sqrt(n()),
            n = n()) |> 
  ungroup() |> 
  arrange(desc(mean_pc))

species_rank_top20 <- species_rank |> 
  filter(n > 1) |> 
  slice_max(mean_pc, n = 20)
```

```{r}
#| label: identify-dominant-species
#| echo: true
# identify top 5 dominant species from species_rank
dominant_species <- species_rank |> 
  slice_max(mean_pc, n = 5) |> 
  pull(psoc)

# summarize dominant species cover by year — raw mean percent cover
dominant_by_year <- saltmarsh_plants |> 
  # label dominant species; group rest as Other
  mutate(psoc_grouped = if_else(psoc %in% dominant_species, psoc, "Other")) |> 
  # filter to only dominant species
  filter(psoc_grouped != "Other") |>
  # get transect-level totals per species per year first
  group_by(year, transect_name, psoc_grouped, cover_category) |> 
  summarize(sum_pc = sum(mean_pc, na.rm = TRUE)) |> 
  ungroup() |> 
  # then average across transects within each year and species
  group_by(year, psoc_grouped, cover_category) |> 
  summarize(mean_pc = mean(sum_pc, na.rm = TRUE)) |> 
  ungroup()
```

```{r}
#| label: summarize-native-nonnative
#| echo: true
native_nonnative <- saltmarsh_plants |> 
  group_by(year, cover_category) |> 
  summarize(mean_pc = mean(mean_pc, na.rm = TRUE)) |> 
  ungroup() |> 
  group_by(year) |> 
  mutate(prop = mean_pc / sum(mean_pc)) |> 
  ungroup()
```

```{r}
#| label: summarize-top-nonnative-species
#| echo: true

top_nonnative_species <- saltmarsh_plants |>
  filter(cover_category == "NON-NATIVE COVER") |>
  group_by(psoc) |>
  summarize(mean_pc = mean(mean_pc, na.rm = TRUE),
            .groups = "drop") |>
  slice_max(mean_pc, n = 5) |>
  pull(psoc)

#Summarizing top non-native species by year
top_nonnative_by_year <- saltmarsh_plants |>
  filter(cover_category == "NON-NATIVE COVER",
         psoc %in% top_nonnative_species) |>
  group_by(year, transect_name, psoc) |>
  summarize(sum_pc = sum(mean_pc, na.rm = TRUE),
            .groups = "drop") |>
  group_by(year, psoc) |>
  summarize(mean_pc = mean(sum_pc, na.rm = TRUE),
            .groups = "drop")
```

```{r}
#| label: summarize-total-cover
#| echo: true
total_cover <- saltmarsh_plants |> 
  group_by(year, transect_name) |> 
  summarize(total_pc = sum(mean_pc, na.rm = TRUE)) |> 
  ungroup() |> 
  group_by(year) |> 
  summarize(mean_total = mean(total_pc, na.rm = TRUE),
            se_total = sd(total_pc, na.rm = TRUE) / sqrt(n())) |> 
  ungroup()
```

# Visualizations directly relevant to answering questions

The following four figures directly address our project questions by connecting plant identity, cover, and origin through time. Figure 1 focuses on the five dominant plant species overall, Figure 2 compares the proportional contribution of native and non-native cover, Figure 3 examines which non-native species contribute most to non-native cover, and Figure 4 ranks the top species by mean cover to justify our dominant-species cutoff.

**Figure 1: Dominant species percent cover across survey years**

```{r}
#| label: fig-dominant-species
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 8
#| fig-cap: "Mean percent cover of the five dominant salt marsh plant species at NCOS across survey years, averaged across transects within each year. Dominant species were identified as the top five by mean percent cover across all survey years. Bar height reflects total mean cover of the five dominant species and varies across years. Green shades indicate native species; coral indicates non-native species."

fig1 <- ggplot(data = dominant_by_year,
               aes(x = year,
                   y = mean_pc,
                   fill = psoc_grouped)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = c(
  "Salicornia_pacifica"     = "#0D5E45",
  "Distichlis_spicata"      = "#1D9E75",
  "Atriplex_lentiformis"    = "#5DCAA5",
  "Eleocharis_macrostachya" = "#A8DBC9",
  "Raphanus_sativus"        = "#D85A30"
)) +
  scale_x_continuous(breaks = seq(2018, 2025, 1)) +
  labs(x = "Survey year",
       y = "Mean percent cover (%)",
       fill = "Species") +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.text = element_text(face = "italic"),
        legend.title = element_text(face = "bold"))

fig1
```

**Figure 2: Proportional native and non-native cover across survey years**

```{r}
#| label: fig-native-nonnative
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 8
#| fig-cap: "Proportional contribution of native and non-native vegetation cover in salt marsh habitat at NCOS across survey years. Each bar sums to 1. Native cover is shown in green; non-native cover in coral."

fig2 <- ggplot(data = native_nonnative,
               aes(x = year,
                   y = prop,
                   fill = cover_category)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = c("NATIVE COVER" = "#1D9E75",
                                "NON-NATIVE COVER" = "#D85A30")) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Survey year",
       y = "Proportion of vegetation cover",
       fill = "Cover category") +
  theme_minimal() +
  theme(legend.position = "bottom")

fig2
```

**Figure 3: Dominant non-native species percent cover across survey years**

```{r}
#| label: fig-3-top-nonnative-species
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 10
#| fig-cap: "Mean percent cover of the five most common non-native plant species in the filtered salt marsh/brackish marsh dataset across survey years, averaged across transects within each year. Bar height reflects the summed mean cover of these non-native species in each year."

#Creating a red/coral color palette for the top non-native species
nonnative_palette <- c("#7F1D1D", "#B8322A", "#D85A30", "#E9825A", "#F3B49C")

#Visualizing top non-native species over time
fig_3 <- ggplot(data = top_nonnative_by_year,
                        aes(x = year,
                            y = mean_pc,
                            fill = psoc)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = nonnative_palette) +
  scale_x_continuous(breaks = seq(2018, 2025, 1)) +
  labs(x = "Survey year",
       y = "Mean percent cover (%)",
       fill = "Non-native species") +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.text = element_text(face = "italic"),
        legend.title = element_text(face = "bold"))

fig_3
```

**Figure 4: Species rank plot**

```{r}
#| label: fig-species-rank
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 8
#| fig-cap: "Mean percent cover (± 1 SE) of the top 20 plant species recorded in salt marsh habitat at NCOS, averaged across transects and survey years and ranked from highest to lowest mean cover. Points represent means and horizontal bars represent ± 1 standard error. Green indicates native species; coral indicates non-native species. Dashed line marks the cutoff for the five dominant species visualized in figure 1."

# fig3 — pointrange instead of col
cutoff_x <- mean(c(species_rank$mean_pc[5], species_rank$mean_pc[6]))

fig4 <- ggplot(data = species_rank_top20,
               aes(x = mean_pc,
                   y = reorder(psoc, mean_pc),
                   color = cover_category)) +
  geom_pointrange(aes(xmin = mean_pc - se_pc,
                      xmax = mean_pc + se_pc)) +
  geom_vline(xintercept = cutoff_x,
             linetype = "dashed",
             color = "gray50") +
  scale_color_manual(values = c("NATIVE COVER" = "#1D9E75",
                                "NON-NATIVE COVER" = "#D85A30")) +
  labs(x = "Mean percent cover (± 1 SE)",
       y = "Species",
       color = "Cover category") +
  theme_minimal() +
  theme(axis.text.y = element_text(face = "italic", size = 9),
        legend.position = "bottom",
        legend.title = element_text(face = "bold"))

fig4
```

# Other exploratory visualizations

**Figure 5: Total vegetation cover across survey years**

```{r}
#| label: fig-total-cover
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 8
#| fig-cap: "Total mean percent vegetation cover (native + non-native combined) in salt marsh habitat at NCOS across survey years. Shaded band represents ± 1 standard error around the mean."

fig5 <- ggplot(data = total_cover,
               aes(x = year,
                   y = mean_total)) +
  geom_ribbon(aes(ymin = mean_total - se_total,
                  ymax = mean_total + se_total),
              fill = "#1D9E75",
              alpha = 0.2) +
  geom_line(color = "#1D9E75",
            linewidth = 1) +
  geom_point(color = "#1D9E75",
             size = 3) +
  labs(x = "Survey year",
       y = "Total mean percent cover (%)") +
  theme_minimal()

fig5
```

# Plan for elective (if completing)

For our advanced elective, we plan to create an educational magazine about the salt marsh plants of North Campus Open Space. The magazine will translate our vegetation analysis into a more accessible and visual format for younger students and community members. Our goal is to introduce readers to the dominant plant species that characterize salt marsh habitat at NCOS while also explaining why salt marshes matter for restoration, wildlife habitat, and coastal resilience.

## Elective concept 

The magazine will be inspired by educational nature magazines such as *National Geographic Kids*. It will include short species spotlight pages for the dominant salt marsh plants identified in our analysis, a simple explanation of native and non-native vegetation, and a data feature that communicates one of our main project figures in a more approachable way. We want the magazine to feel informative, visual, and welcoming for students who may not have much background in ecology.

## Draft sections (include some description/image of drafts)

Our magazine sections will include a short “Meet the Marsh” introduction that explains what salt marsh habitat is and why it matters at NCOS. We also plan to include a spotlight page for the dominant salt marsh plants identified in our analysis, featuring drawings, the species’ native or non-native status, and a brief description of its role in the marsh community.

Our magazine will also include a “Native, Non-native, or Invasive?” page and a data visualization that translates one of our figures into a more accessible visual. We also want to include a final “How to Get Involved” page that connects younger students back to NCOS by explaining how they can visit, learn more, or care for restored coastal habitats in their own communities.

## Plan for the next two weeks 

Over the next two weeks, we will use our preliminary analysis to select the species featured in the magazine. We will also continue revising our figures so the main results are clear enough to translate into a more visual and accessible format for younger students and community readers.

By Week 9, we hope to draft the species spotlight pages, begin the species drawings, and write the native/non-native explanation. We will also begin designing the magazine layout in Canva and deciding on a visual style that fits the salt marsh theme. By Week 10, we plan to complete the full magazine draft, revise the writing for clarity, and polish the layout so the final product is cohesive, educational, and visually engaging.

# Next steps

Our next steps are to refine the statistical analyses and continue revising the figures so our interpretations are clear and well-supported. For the final project, we will add more complete citations, connect our results more directly to NCOS restoration goals, and continue developing the advanced elective magazine alongside the data analysis.

Before submission, we will make sure the GitHub repository includes the rendered PDF, the QMD, the data files, the bibliography file, and a README that clearly links to the rendered document.

# References
