---
title: "Salt Marsh Vegetation at NCOS"
author: "Macey Hartmann and Erick Morales Oyola"
format: pdf
---
  
# Guiding questions

- Which dominant plant species characterize salt marsh habitat at NCOS?
- How has the mean percent cover of the dominant salt marsh species changed across survey years?
- How has the relative contribution of native and non-native vegetation changed across survey years?


# 1. Set up

## Packages

```{r}
#| label: packages

# general use
library(tidyverse)

# file organization
library(here)

# data visualization
library(patchwork)
library(naniar)
```



## Data

```{r}
#| label: read-data

veg <- read_csv(here("data", "veg.csv"))

metadata <- read_csv(here("data", "vp_veg_metadata.csv"))
```

# 2. Cleaning and wrangling

## Explore missingness

```{r}
#| label: missingness

gg_miss_var(veg)
```

## Filter to salt marsh habitat and late summer/early fall surveys

```{r}
#| label: filter-saltmarsh

saltmarsh <- veg |> 
  # filter to salt marsh habitat only
  filter(str_detect(habitat, "Salt Marsh") | str_detect(habitat, "Brackish Marsh")) |>
  # parse date column and filter to August and September
  mutate(date = as.Date(date),
         month = month(date)) |> 
  filter(month %in% c(8, 9))
```

## Fill implicit NAs with 0

```{r}
#| label: explicit-zeros
saltmarsh_complete <- saltmarsh |> 
  complete(year, transect_name, psoc, transect_distance,
           fill = list(percent_cover = 0)) |> 
  group_by(year, transect_name, psoc, cover_category) |> 
  summarize(sum_pc = sum(percent_cover, na.rm = TRUE)) |> 
  ungroup() |>
  unite("year_pool", year, transect_name, remove = FALSE) |> 
  left_join(metadata, by = "year_pool") |>
  select(-year.y, -transect_name.y) |>
  rename(year = year.x,
         transect_name = transect_name.x) |>
  mutate(mean_pc = sum_pc / num_quad)
```

## Filter to plant cover categories only

```{r}
#| label: filter-plant-cover

saltmarsh_plants <- saltmarsh_complete |> 
  filter(cover_category %in% c("NATIVE COVER", "NON-NATIVE COVER"))
```


# 3. Summarizing for each figure

## Figure 3: Species rank (averaged across all years)

```{r}
#| label: summarize-species-rank

species_rank <- saltmarsh_plants |> 
  group_by(psoc, cover_category) |> 
  summarize(mean_pc = mean(mean_pc, na.rm = TRUE)) |> 
  ungroup() |> 
  arrange(desc(mean_pc))

# print to identify dominant species
species_rank
```


## Figure 1: Dominant species by year

```{r}
#| label: identify-dominant-species

# identify top 5 dominant species from species_rank
dominant_species <- species_rank |> 
  slice_max(mean_pc, n = 5) |> 
  pull(psoc)

# summarize dominant species cover by year
dominant_by_year <- saltmarsh_plants |> 
  mutate(psoc_grouped = if_else(psoc %in% dominant_species, psoc, "Other")) |> 
  group_by(year, psoc_grouped, cover_category) |> 
  summarize(mean_pc = mean(mean_pc, na.rm = TRUE)) |> 
  ungroup()
```

## Figure 2: Native/non-native proportions by year

```{r}
#| label: summarize-native-nonnative

native_nonnative <- saltmarsh_plants |> 
  group_by(year, cover_category) |> 
  summarize(mean_pc = mean(mean_pc, na.rm = TRUE)) |> 
  ungroup() |> 
  group_by(year) |> 
  mutate(prop = mean_pc / sum(mean_pc)) |> 
  ungroup()
```

## Figure 4: Total vegetation cover by year

```{r}
#| label: summarize-total-cover

total_cover <- saltmarsh_plants |> 
  group_by(year, transect_name) |> 
  summarize(total_pc = sum(mean_pc, na.rm = TRUE)) |> 
  ungroup() |> 
  group_by(year) |> 
  summarize(mean_total = mean(total_pc, na.rm = TRUE),
            se_total = sd(total_pc, na.rm = TRUE) / sqrt(n())) |> 
  ungroup()
```

# 4. Visualizations

## Figure 1: Dominant species percent cover across survey years

```{r}
#| label: fig-dominant-species
#| fig-cap: "Mean percent cover of dominant salt marsh plant species at NCOS across survey years. The top 5 species by mean cover across all years are shown; remaining species are grouped as Other. Colors indicate native (green) or non-native (coral) status."

fig1 <- ggplot(data = dominant_by_year,
               aes(x = year,
                   y = mean_pc,
                   fill = psoc_grouped)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = c(
    # placeholders — update with actual dominant species names after running
    "Other" = "#B4B2A9"
  )) +
  labs(x = "Survey year",
       y = "Mean percent cover (%)",
       fill = "Species") +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.text = element_text(face = "italic"))

fig1
```

## Figure 2: Proportional native vs. non-native cover across survey years

```{r}
#| label: fig-native-nonnative
#| fig-cap: "Proportional contribution of native and non-native vegetation cover in salt marsh habitat at NCOS across survey years. Each bar sums to 1. Native cover is shown in green; non-native cover in coral."

fig2 <- ggplot(data = native_nonnative,
               aes(x = year,
                   y = prop,
                   fill = cover_category)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = c("NATIVE COVER" = "#1D9E75",
                                "NON-NATIVE COVER" = "#D85A30")) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Survey year",
       y = "Proportion of vegetation cover",
       fill = "Cover category") +
  theme_minimal() +
  theme(legend.position = "bottom")

fig2
```

## Figure 3: Species rank plot

```{r}
#| label: fig-species-rank
#| fig-cap: "Mean percent cover of plant species recorded in salt marsh habitat at NCOS, averaged across all survey years and ranked from highest to lowest. Green bars indicate native species; coral bars indicate non-native species. Dashed line marks the dominant species cutoff used in figure 1."

fig3 <- ggplot(data = species_rank,
               aes(x = mean_pc,
                   y = reorder(psoc, mean_pc),
                   fill = cover_category)) +
  geom_col() +
  geom_vline(xintercept = species_rank$mean_pc[5],
             linetype = "dashed",
             color = "gray50") +
  scale_fill_manual(values = c("NATIVE COVER" = "#1D9E75",
                                "NON-NATIVE COVER" = "#D85A30")) +
  labs(x = "Mean percent cover (%)",
       y = "Species",
       fill = "Cover category") +
  theme_minimal() +
  theme(axis.text.y = element_text(face = "italic"),
        legend.position = "bottom")

fig3
```

## Figure 4: Total vegetation cover across survey years

```{r}
#| label: fig-total-cover
#| fig-cap: "Total mean percent vegetation cover (native + non-native combined) in salt marsh habitat at NCOS across survey years. Shaded band represents ± 1 standard error around the mean."

fig4 <- ggplot(data = total_cover,
               aes(x = year,
                   y = mean_total)) +
  geom_ribbon(aes(ymin = mean_total - se_total,
                  ymax = mean_total + se_total),
              fill = "#1D9E75",
              alpha = 0.2) +
  geom_line(color = "#1D9E75",
            linewidth = 1) +
  geom_point(color = "#1D9E75",
             size = 3) +
  labs(x = "Survey year",
       y = "Total mean percent cover (%)") +
  theme_minimal()

fig4
```


