---
title: "Salt Marsh Vegetation at NCOS"
author: "Macey Hartmann and Erick Morales Oyola"
format: pdf
---
  
# Guiding questions

- Which dominant plant species characterize salt marsh habitat at NCOS? (figure 3, showing the full species rank and identifying which species have the highest mean cover overall)
- How has the mean percent cover of the dominant salt marsh species changed across survey years? (figure 1, showing top species' cover through time)
- How has the relative contribution of native and non-native vegetation changed across survey years? (figure 2, showing native vs. non-native cover through time)


# 1. Set up

## Packages

```{r}
#| label: packages

# general use
library(tidyverse)

# file organization
library(here)

# data visualization
library(patchwork)
library(naniar)
```



## Data

```{r}
#| label: read-data

veg <- read_csv(here("data", "veg.csv"))

metadata <- read_csv(here("data", "vp_veg_metadata.csv"))
```

# 2. Cleaning and wrangling

## Explore missingness

```{r}
#| label: missingness

gg_miss_var(veg)
```

## Filter to salt marsh habitat and late summer/early fall surveys

```{r}
#| label: filter-saltmarsh

saltmarsh <- veg |> 
  # filter to salt marsh habitat only
  filter(str_detect(habitat, "Salt Marsh") | str_detect(habitat, "Brackish Marsh")) |>
  # parse date column and filter to August and September
  mutate(date = as.Date(date),
         month = month(date)) |> 
  filter(month %in% c(8, 9))
```

## Fill implicit NAs with 0

```{r}
#| label: explicit-zeros
saltmarsh_complete <- saltmarsh |> 
  complete(year, transect_name, psoc, transect_distance,
           fill = list(percent_cover = 0)) |> 
  group_by(year, transect_name, psoc, cover_category) |> 
  summarize(sum_pc = sum(percent_cover, na.rm = TRUE)) |> 
  ungroup() |>
  unite("year_pool", year, transect_name, remove = FALSE) |> 
  left_join(metadata, by = "year_pool") |>
  select(-year.y, -transect_name.y) |>
  rename(year = year.x,
         transect_name = transect_name.x) |>
  mutate(mean_pc = sum_pc / num_quad)
```

## Filter to plant cover categories only

```{r}
#| label: filter-plant-cover

saltmarsh_plants <- saltmarsh_complete |> 
  filter(cover_category %in% c("NATIVE COVER", "NON-NATIVE COVER"))
```


# 3. Summarizing for each figure

## Figure 3: Species rank (averaged across all years)

```{r}
#| label: summarize-species-rank

# summarize-species-rank — keep SE this time
species_rank <- saltmarsh_plants |> 
  group_by(psoc, cover_category) |> 
  summarize(mean_pc = mean(mean_pc, na.rm = TRUE),
            se_pc = sd(mean_pc, na.rm = TRUE) / sqrt(n()),
            n = n()) |> 
  ungroup() |> 
  arrange(desc(mean_pc))

species_rank_top20 <- species_rank |> 
  filter(n > 1) |> 
  slice_max(mean_pc, n = 20)
```


## Figure 1: Dominant species by year

```{r}
#| label: identify-dominant-species

# identify top 5 dominant species from species_rank
dominant_species <- species_rank |> 
  slice_max(mean_pc, n = 5) |> 
  pull(psoc)

# summarize dominant species cover by year — raw mean percent cover
dominant_by_year <- saltmarsh_plants |> 
  # label dominant species; group rest as Other
  mutate(psoc_grouped = if_else(psoc %in% dominant_species, psoc, "Other")) |> 
  # filter to only dominant species
  filter(psoc_grouped != "Other") |>
  # get transect-level totals per species per year first
  group_by(year, transect_name, psoc_grouped, cover_category) |> 
  summarize(sum_pc = sum(mean_pc, na.rm = TRUE)) |> 
  ungroup() |> 
  # then average across transects within each year and species
  group_by(year, psoc_grouped, cover_category) |> 
  summarize(mean_pc = mean(sum_pc, na.rm = TRUE)) |> 
  ungroup()
```

## Figure 2: Native/non-native proportions by year

```{r}
#| label: summarize-native-nonnative

native_nonnative <- saltmarsh_plants |> 
  group_by(year, cover_category) |> 
  summarize(mean_pc = mean(mean_pc, na.rm = TRUE)) |> 
  ungroup() |> 
  group_by(year) |> 
  mutate(prop = mean_pc / sum(mean_pc)) |> 
  ungroup()
```

## Figure 4: Total vegetation cover by year

```{r}
#| label: summarize-total-cover

total_cover <- saltmarsh_plants |> 
  group_by(year, transect_name) |> 
  summarize(total_pc = sum(mean_pc, na.rm = TRUE)) |> 
  ungroup() |> 
  group_by(year) |> 
  summarize(mean_total = mean(total_pc, na.rm = TRUE),
            se_total = sd(total_pc, na.rm = TRUE) / sqrt(n())) |> 
  ungroup()
```

# 4. Visualizations

## Figure 1: Dominant species percent cover across survey years

```{r}
#| label: fig-dominant-species
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 8
#| fig-cap: "Mean percent cover of the five dominant salt marsh plant species at NCOS across survey years, averaged across transects within each year. Dominant species were identified as the top five by mean percent cover across all survey years. Bar height reflects total mean cover of the five dominant species and varies across years. Green shades indicate native species; coral indicates non-native species."

fig1 <- ggplot(data = dominant_by_year,
               aes(x = year,
                   y = mean_pc,
                   fill = psoc_grouped)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = c(
  "Salicornia_pacifica"     = "#0D5E45",
  "Distichlis_spicata"      = "#1D9E75",
  "Atriplex_lentiformis"    = "#5DCAA5",
  "Eleocharis_macrostachya" = "#A8DBC9",
  "Raphanus_sativus"        = "#D85A30"
)) +
  scale_x_continuous(breaks = seq(2018, 2025, 1)) +
  labs(x = "Survey year",
       y = "Mean percent cover (%)",
       fill = "Species") +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.text = element_text(face = "italic"),
        legend.title = element_text(face = "bold"))

fig1
```

## Figure 2: Proportional native vs. non-native cover across survey years

```{r}
#| label: fig-native-nonnative
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 8
#| fig-cap: "Proportional contribution of native and non-native vegetation cover in salt marsh habitat at NCOS across survey years. Each bar sums to 1. Native cover is shown in green; non-native cover in coral."

fig2 <- ggplot(data = native_nonnative,
               aes(x = year,
                   y = prop,
                   fill = cover_category)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = c("NATIVE COVER" = "#1D9E75",
                                "NON-NATIVE COVER" = "#D85A30")) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Survey year",
       y = "Proportion of vegetation cover",
       fill = "Cover category") +
  theme_minimal() +
  theme(legend.position = "bottom")

fig2
```

## Figure 3: Species rank plot

```{r}
#| label: fig-species-rank
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 8
#| fig-cap: "Mean percent cover (± 1 SE) of the top 20 plant species recorded in salt marsh habitat at NCOS, averaged across transects and survey years and ranked from highest to lowest mean cover. Points represent means and horizontal bars represent ± 1 standard error. Green indicates native species; coral indicates non-native species. Dashed line marks the cutoff for the five dominant species visualized in figure 1."

# fig3 — pointrange instead of col
cutoff_x <- mean(c(species_rank$mean_pc[5], species_rank$mean_pc[6]))

fig3 <- ggplot(data = species_rank_top20,
               aes(x = mean_pc,
                   y = reorder(psoc, mean_pc),
                   color = cover_category)) +
  geom_pointrange(aes(xmin = mean_pc - se_pc,
                      xmax = mean_pc + se_pc)) +
  geom_vline(xintercept = cutoff_x,
             linetype = "dashed",
             color = "gray50") +
  scale_color_manual(values = c("NATIVE COVER" = "#1D9E75",
                                "NON-NATIVE COVER" = "#D85A30")) +
  labs(x = "Mean percent cover (± 1 SE)",
       y = "Species",
       color = "Cover category") +
  theme_minimal() +
  theme(axis.text.y = element_text(face = "italic", size = 9),
        legend.position = "bottom",
        legend.title = element_text(face = "bold"))

fig3
```

## Figure 4: Total vegetation cover across survey years

```{r}
#| label: fig-total-cover
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 8
#| fig-cap: "Total mean percent vegetation cover (native + non-native combined) in salt marsh habitat at NCOS across survey years. Shaded band represents ± 1 standard error around the mean."

fig4 <- ggplot(data = total_cover,
               aes(x = year,
                   y = mean_total)) +
  geom_ribbon(aes(ymin = mean_total - se_total,
                  ymax = mean_total + se_total),
              fill = "#1D9E75",
              alpha = 0.2) +
  geom_line(color = "#1D9E75",
            linewidth = 1) +
  geom_point(color = "#1D9E75",
             size = 3) +
  labs(x = "Survey year",
       y = "Total mean percent cover (%)") +
  theme_minimal()

fig4
```


