---
title: "Timeline Check In"
author: "Macey Hartmann and Erick Morales Oyola"
date: "05-18-2026"
format: 
  pdf:
    include-in-header: 
       text: |
         \usepackage{fvextra}
         \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
         \DefineVerbatimEnvironment{OutputCode}{Verbatim}{breaklines,commandchars=\\\{\}}
editor: source
execute: 
  message: false
  warning: false
---

```{r}
#| label: packages
#| echo: false

# general use
library(tidyverse)

# file organization
library(here)

# data visualization
library(patchwork)
library(naniar)
```


```{r}
#| label: read-data
#| echo: false

veg <- read_csv(here("data", "veg.csv"))

metadata <- read_csv(here("data", "vp_veg_metadata.csv"))
```


```{r}
#| label: missingness
#| echo: false
#| include: false
gg_miss_var(veg)
```


```{r}
#| label: filter-saltmarsh
#| echo: false
saltmarsh <- veg |> 
  # filter to salt marsh habitat only
  filter(str_detect(habitat, "Salt Marsh") | str_detect(habitat, "Brackish Marsh")) |>
  # parse date column and filter to August and September
  mutate(date = as.Date(date),
         month = month(date)) |> 
  filter(month %in% c(8, 9))
```


```{r}
#| label: explicit-zeros
#| echo: false
saltmarsh_complete <- saltmarsh |> 
  complete(year, transect_name, psoc, transect_distance,
           fill = list(percent_cover = 0)) |> 
  group_by(year, transect_name, psoc, cover_category) |> 
  summarize(sum_pc = sum(percent_cover, na.rm = TRUE)) |> 
  ungroup() |>
  unite("year_pool", year, transect_name, remove = FALSE) |> 
  left_join(metadata, by = "year_pool") |>
  select(-year.y, -transect_name.y) |>
  rename(year = year.x,
         transect_name = transect_name.x) |>
  mutate(mean_pc = sum_pc / num_quad)
```


```{r}
#| label: filter-plant-cover
#| echo: false
saltmarsh_plants <- saltmarsh_complete |> 
  filter(cover_category %in% c("NATIVE COVER", "NON-NATIVE COVER"))
```


```{r}
#| label: summarize-species-rank
#| echo: false
# summarize-species-rank — keep SE this time
species_rank <- saltmarsh_plants |> 
  group_by(psoc, cover_category) |> 
  summarize(mean_pc = mean(mean_pc, na.rm = TRUE),
            se_pc = sd(mean_pc, na.rm = TRUE) / sqrt(n()),
            n = n()) |> 
  ungroup() |> 
  arrange(desc(mean_pc))

species_rank_top20 <- species_rank |> 
  filter(n > 1) |> 
  slice_max(mean_pc, n = 20)
```

```{r}
#| label: identify-dominant-species
#| echo: false
# identify top 5 dominant species from species_rank
dominant_species <- species_rank |> 
  slice_max(mean_pc, n = 5) |> 
  pull(psoc)

# summarize dominant species cover by year — raw mean percent cover
dominant_by_year <- saltmarsh_plants |> 
  # label dominant species; group rest as Other
  mutate(psoc_grouped = if_else(psoc %in% dominant_species, psoc, "Other")) |> 
  # filter to only dominant species
  filter(psoc_grouped != "Other") |>
  # get transect-level totals per species per year first
  group_by(year, transect_name, psoc_grouped, cover_category) |> 
  summarize(sum_pc = sum(mean_pc, na.rm = TRUE)) |> 
  ungroup() |> 
  # then average across transects within each year and species
  group_by(year, psoc_grouped, cover_category) |> 
  summarize(mean_pc = mean(sum_pc, na.rm = TRUE)) |> 
  ungroup()
```


```{r}
#| label: summarize-native-nonnative
#| echo: false
native_nonnative_raw <- saltmarsh_plants |> 
  group_by(year, transect_name, cover_category) |> 
  summarize(total_pc = sum(mean_pc, na.rm = TRUE)) |> 
  ungroup() |> 
  group_by(year, cover_category) |> 
  summarize(mean_cover = mean(total_pc, na.rm = TRUE),
            se_cover = sd(total_pc, na.rm = TRUE) / sqrt(n())) |> 
  ungroup()
```


```{r}
#| label: summarize-total-cover
#| echo: false
total_cover <- saltmarsh_plants |> 
  group_by(year, transect_name) |> 
  summarize(total_pc = sum(mean_pc, na.rm = TRUE)) |> 
  ungroup() |> 
  group_by(year) |> 
  summarize(mean_total = mean(total_pc, na.rm = TRUE),
            se_total = sd(total_pc, na.rm = TRUE) / sqrt(n())) |> 
  ungroup()
```

# Ideas for analysis and background
## Should include some citations (not all) 
## Should include ideas for analyses to answer questions and some preliminary code to run those analyses

# Visualizations directly relevant to answering questions

**Figure 1: Dominant species percent cover across survey years**

```{r}
#| label: fig-dominant-species
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 8
#| fig-cap: "Mean percent cover of the five dominant salt marsh plant species at NCOS across survey years, averaged across transects within each year. Dominant species were identified as the top five by mean percent cover across all survey years. Bar height reflects total mean cover of the five dominant species and varies across years. Green shades indicate native species; coral indicates non-native species."

fig1 <- ggplot(data = dominant_by_year,
               aes(x = year,
                   y = mean_pc,
                   fill = psoc_grouped)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = c(
  "Salicornia_pacifica"     = "#0D5E45",
  "Distichlis_spicata"      = "#1D9E75",
  "Atriplex_lentiformis"    = "#5DCAA5",
  "Eleocharis_macrostachya" = "#A8DBC9",
  "Raphanus_sativus"        = "#D85A30"
)) +
  scale_x_continuous(breaks = seq(2018, 2025, 1)) +
  labs(x = "Survey year",
       y = "Mean percent cover (%)",
       fill = "Species") +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.text = element_text(face = "italic"),
        legend.title = element_text(face = "bold"))

fig1
```

The original figure was unclear about what the y-axis represented. The wrangling approach was revised so that rather than averaging directly across all observations, we first calculated transect-level totals per species per year and then averaged those totals across transects. This ensures the y-axis represents true mean percent cover averaged across transects within each year, and bar height now varies across years. The "Other" category was also removed entirely so the figure shows only the five dominant species.

**Figure 2: Proportional native vs. non-native cover across survey years**

```{r}
#| label: fig-native-nonnative
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 8
#| fig-cap: "Mean percent cover (± 1 SE) of native and non-native vegetation in salt marsh habitat at NCOS across survey years. Points represent means and shaded bands represent ± 1 standard error. Native cover is shown in green; non-native cover in coral."

fig2 <- ggplot(data = native_nonnative_raw,
               aes(x = year,
                   y = mean_cover,
                   color = cover_category,
                   fill = cover_category)) +
  geom_ribbon(aes(ymin = mean_cover - se_cover,
                  ymax = mean_cover + se_cover),
              alpha = 0.2,
              color = NA) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  scale_color_manual(values = c("NATIVE COVER" = "#1D9E75",
                                "NON-NATIVE COVER" = "#D85A30")) +
  scale_fill_manual(values = c("NATIVE COVER" = "#1D9E75",
                               "NON-NATIVE COVER" = "#D85A30")) +
  scale_x_continuous(breaks = seq(2018, 2025, 1)) +
  labs(x = "Survey year",
       y = "Mean percent cover (± 1 SE)",
       color = "Cover category",
       fill = "Cover category") +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.title = element_text(face = "bold"))

fig2
```

The original proportional stacked bar was nearly identical to panel (b) of figure 9 in the NCOS monitoring report, making it redundant. We replaced it with a line plot showing raw mean percent cover of native and non-native vegetation over time, with ± 1 standard error ribbons around each line. This is distinct from the monitoring report because it uses a continuous line geometry that emphasizes temporal trend, places both cover categories on the same axis for direct visual comparison, and includes a measure of uncertainty that the monitoring report's bar charts do not show.

**Figure 3: Species rank plot**

```{r}
#| label: fig-species-rank
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 8
#| fig-cap: "Mean percent cover (± 1 SE) of the top 20 plant species recorded in salt marsh habitat at NCOS, averaged across transects and survey years and ranked from highest to lowest mean cover. Points represent means and horizontal bars represent ± 1 standard error. Green indicates native species; coral indicates non-native species. Dashed line marks the cutoff for the five dominant species visualized in figure 1."

# fig3 — pointrange instead of col
cutoff_x <- mean(c(species_rank$mean_pc[5], species_rank$mean_pc[6]))

fig3 <- ggplot(data = species_rank_top20,
               aes(x = mean_pc,
                   y = reorder(psoc, mean_pc),
                   color = cover_category)) +
  geom_pointrange(aes(xmin = mean_pc - se_pc,
                      xmax = mean_pc + se_pc)) +
  geom_vline(xintercept = cutoff_x,
             linetype = "dashed",
             color = "gray50") +
  scale_color_manual(values = c("NATIVE COVER" = "#1D9E75",
                                "NON-NATIVE COVER" = "#D85A30")) +
  labs(x = "Mean percent cover (± 1 SE)",
       y = "Species",
       color = "Cover category") +
  theme_minimal() +
  theme(axis.text.y = element_text(face = "italic", size = 9),
        legend.position = "bottom",
        legend.title = element_text(face = "bold"))

fig3
```

The original figure used geom_col to show only mean percent cover as a horizontal bar with no representation of spread. Following the instructor's suggestion, we switched to geom_pointrange where points represent species means and horizontal bars represent ± 1 standard error, calculated across transect-year observations. We also filtered to species with more than one observation before plotting so that SE could actually be calculated, and limited the figure to the top 20 species by mean cover to keep the y-axis readable.

# Other exploratory visualizations

**Figure 4: Total vegetation cover across survey years**

```{r}
#| label: fig-total-cover
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 8
#| fig-cap: "Total mean percent vegetation cover (native + non-native combined) in salt marsh habitat at NCOS across survey years. Shaded band represents ± 1 standard error around the mean."

fig4 <- ggplot(data = total_cover,
               aes(x = year,
                   y = mean_total)) +
  geom_ribbon(aes(ymin = mean_total - se_total,
                  ymax = mean_total + se_total),
              fill = "#1D9E75",
              alpha = 0.2) +
  geom_line(color = "#1D9E75",
            linewidth = 1) +
  geom_point(color = "#1D9E75",
             size = 3) +
  labs(x = "Survey year",
       y = "Total mean percent cover (%)") +
  theme_minimal()

fig4
```


# Plan for elective (if completing)

## Should include some description/image of drafts

## Should include plan for making progress in the next two weeks

For our advanced elective, we are creating a National Geographic-inspired educational magazine about the salt marsh plants of the North Campus Open Space (NCOS). The magazine is designed for a fourth grade audience and will introduce readers to the dominant plant species that characterize salt marsh habitat at NCOS. Sections will include species spotlight pages profiling the dominant salt marsh plants with illustrated profiles, fun facts, and native/non-native status; a "Meet the Marsh" illustrated map of NCOS showing habitat zones; a data feature communicating our project findings; a "Native vs. Non-Native" explanation page; a salt marsh vocabulary word search; a "What Plant Are You?" personality flowchart; and a how-to-get-involved page describing how to visit and learn more about NCOS. The magazine will be designed in Canva, printed in color at the dorm print facilities, and delivered to Ms. McCarter's fourth grade classroom with one copy per table group (approximately 3–4 copies total).

We have finalized the magazine structure, section outline, and visual style. Based on our preliminary data exploration, we have identified the dominant salt marsh species to feature: Salicornia pacifica, Distichlis spicata, Atriplex lentiformis, and Raphanus sativus. For images, we will use Creative Commons licensed photographs from iNaturalist, Wikimedia Commons, and the USDA PLANTS Database, supplement with our own photographs taken at NCOS, and create original botanical-style illustrations for species that are difficult to photograph clearly.

By the week 9 check-in, we will have completed all remaining sections including the data feature, word search, flowchart, and how-to-get-involved page, and will have begun design and layout in Canva. By week 10 we will finalize and polish the full magazine for presentation. The completed magazine will be delivered to Ms. McCarter's classroom during finals week.


